---
title: "RProgramming"
author: "tubako"
date: "8/30/2020"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# R Console Input and Evaluation
<br>

### Entering input
```{r Entering input}
x <- 1
print(x)
x
msg <- "hello"

## x <- ##Incomplete expression
```

### Evaluation
```{r Evaluation}
x <- 5 ## nothing happens
x ## auto-printing occurs

print(x) ## explicit printing

## [1] indicates that x is a vector and 5 is the first element.
```

### Printing
```{r Printing}
## The : operator is used to create integer sequences.
x <- 1:20 
x
```

<br>

# Data Types 

### R Objects and Attributes


#### Objects 

R has 5 basic classes of objects

- character  
- numeric (real numbers)  
- integer  
- complex  
- logical (T/F)  

The most basic object is a vector


- A vector can only contain objects of the same class  
- BUT: The one exception is a *list*, which is represented as a vector but can contain objects of different classes (indeed, that's usually why we use them)

Empty vectors can be created with the 
```{r}
vector()
```
function.

#### Numbers

- Numbers in R are generally treated as numeric objects (i.e. double precision real numbers)  
- If you explicitly want an integer, you need to specify the **L** suffix  
- Ex: Entering 1 gives you a numeric object; entering 1**L** explicitly gives you an integer.  
- There is also a special number **Inf** which represents infinity: e.g. 1/0; **Inf** can be used in ordinary calculations; e.g. 1/**Inf** is 0  
- The value **NaN** represents an undefined value ("not a number); e.g. 0/0; **NaN** can also be thought of as a missing value  

#### Attributes

R objects can have attributes

- names, dimensions  
- dimensions (e.g. matrices, arrays)  
- class  
- length  
- other user-defined attributes/metadata  

Attributes of an object can be accessed using the 
```{r}
##attributes()
```
function.


<br>

### Vectors and Lists

#### Creating Vectors

The **c()** function can be used to create vectors of objects.

```{r}
x <- c(0.5, 0.6)       ## numeric
x <- c(TRUE, FALSE)    ## logical
x <- c(T, F)           ## logical
x <- c("a", "b", "c")  ## character
x <- 9:29              ## integer
x <- c(1+0i, 2+4i)     ## complex

```

Using the **vector()** function
```{r}
x <- vector("numeric", length = 10)
x
```
#### Mixing Objects

When different objects are mixed in  a vector, *coercion* occurs so that every element in the vector is of the same class.

```{r}
y <- c(1.7, "a")   ## character
y <- c(TRUE, 2)    ## numeric
y <- c("a", TRUE)  ## character
y
```
#### Explicit Coercion

Objects can be explicitly coerced from one class to another using the as.* functions, if available. 

```{r}
x <- 0:6  
class(x)
as.numeric(x)
as.logical(x)
as.character(x)
```
Nonsensical coercion results in **NA**s.
```{r}
x <- c("a", "b", "c")
as.numeric(x)
as.logical(x)
as.complex(x)
```
#### Lists
```{r}
x <- list(1, "a", TRUE, 1+4i)
x

```

### Matrices

The dimension attribute is itself an integer vector of length 2 (nrow, ncol).

```{r}
m <- matrix(nrow = 2, ncol = 3)
m
dim(m)
attributes(m)
```

Matrices are constructed column wise, so entries can be thought of starting in the "upper left" corner and running down the coloumns.

```{r}
m <- matrix (1:6, nrow= 2, ncol = 3)
m
```

You can also create a matrix by creating the dimension attribute on a vector. 
```{r}
m <- 1:10
m
dim (m) <- c(2, 5)
m
```
### cbind-ing and rbind-ing
Matrices can be created by *column-binding* or *row-binding* with **cbind()** and **rbind()**.

```{r}
x <- 1:3
y <- 10:12
cbind(x, y)
rbind(x, y)
```

### Factors
Factors are used to represent categorical data. There's two types of factor, there is unordered or ordered. One can think of a factor as an integer vector where each integer has a *label*.  

- Factors are treated specially by modeling functions like **lm()** and **glm()**  
- Using factors with labels is better than using using integers because factors are self-describing; having a variable that has values "Male" and "Female" is better than a variable that has values 1 and 2.  

```{r}
x <- factor(c("yes", "yes", "no", "yes", "no"))
x
table(x)
unclass(x)
```
The order of the levels can be set using the **levels** argument to **factor()**. This can be important in linear modeling because the fist level is used as the baseline level.
```{r}
x <- factor(c("yes", "yes", "no", "yes", "no"), level = c("yes", "no"))
x
unclass(x)
```
<br>

### Missing Values

Missing values in R are denoted by either **NA** or **NAN**. **NAN** is used for undefined mathematical operations. And **NA** is used for everything else.

- **is.na()** is used to test objects if they are **NA**  
- **is.nan()** is used to test for **NaN**  
- **NA** values have a class also, so there are integer **NA** character **NA**, etc.  
- A **NaN** value is also **NA** but the converse is not true.

<br>

```{r}
x <- c(1, 2, NA, 10, 3)
is.na(x)
is.nan(x)

x <- c(1, 2, NaN, NA, 4)
is.na(x)
is.nan(x)
```
<br>

### Data Frames

Data frames are used to store tabular data.

- They are represented as a special type of list where every element of the list has to have the same length  
- Each element of the list can be thought of as a column and the length of each element of the list is the number of rows  
- Unlike matrices, data frames can store different classes of objects in each column (just like lists); matrices must have every element be the same class  
- Data frames also have a special atttribute called row.names  
- Data frames are usually created by calling **read.table()** or **read.csv()**  
- Can be converted to a matrix by calling **data.matrix()**

<br>

```{r}
x <- data.frame(foo = 1:4, bar = c(T, T, F, F))
x
nrow(x)
ncol(x)
```
<br>

### Names Attribute

#### Names

R objects can also have names, which is very useful for writing readable code and self-describing objects.

```{r}
x <- 1:3
names(x)
names(x) <- c("foo", "bar", "norf")
x
names(x)
```
Lists can also have names.

```{r}
x <- list(a = 1, b = 2, c = 3)
x
```
And matrices.

```{r}
m <- matrix(1:4, nrow = 2, ncol = 2)
dimnames(m) <- list(c("a", "b"), c("c", "d"))
m
```

